---
title: "Neural Progenitor Cells for Treatment of Spinal Cord Injury"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#1. INSTALLING PACKAGES
#install.packages("ggplot2")
#install.packages("data.table")
#install.packages("RColorBrewer")
#install.packages("cowplot")

#2. ATTACHING PACKAGES
library(ggplot2)
library(data.table)
library(RColorBrewer)
library(cowplot)
library(knitr)
```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#1. IMPORTING DATA
DT = fread("SCI_NPC_saline_overtime_data.csv")

#2. CLEANING COLUMN NAMES
column_name_adjuster <- function(column_name_raw){
    column_name_adjusted <- substr(column_name_raw, 4, nchar(column_name_raw))
    column_name_adjusted <- substr(column_name_adjusted, 0, nchar(column_name_adjusted)-5)
    return(column_name_adjusted)
}

col_name_subset <- names(DT[,!c("SCI", "treatment", "evaluation", "evaluation_weeks")])   
col_name_subset <- unlist(lapply(col_name_subset, function(col_name){column_name_adjuster(col_name)}))
names(DT) <-  c("SCI", "treatment", "evaluation", "evaluation_weeks", col_name_subset)

#3. REMOVING UNNECESSARY DATA
DT <-  DT[!is.na(evaluation_weeks)]
DT[, "evaluation"] = NULL

#4. LONG FORMAT
DT_melt <- melt.data.table(DT, id.vars=c("SCI", "treatment", "evaluation_weeks"))

#5. NORMALIZE TO HEALTHY FOR EACH TARGET SEPARATELY
expression_normalizer <-  function(list_object){
  divisor <-  list_object[treatment=="none", mean(value)]
  list_object[, norm_value:=value/divisor]
  return(list_object)
}

DT_melt <- do.call(rbind, lapply(split(DT_melt, DT_melt[,variable]), function(object){expression_normalizer(object)}))

#6. REMOVING UNNECESSARY DATA & ADJUSTING VARIABLE TYPES
DT_melt[, "SCI"] = NULL
DT_melt <- DT_melt[treatment!="none"]
DT_melt[, "treatment"] <- factor(DT_melt[, treatment])
```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
################################################################### STATISTICAL ANALYSIS ##############################################################
```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#1. REMOVING OUTLIERS
#Range based on histogram of norm_values. IL-7 removed due to multiple outliers. 
DT_melt <- DT_melt[norm_value<20 & variable!="IL-7"]
DT_melt[,"variable"] <- factor(DT_melt[,variable])

#2. Summarizing data per treatment, time point and target
DT_summary <- DT_melt[, .(norm_value = mean(norm_value), norm_value_sd = sd(norm_value), n=.N), by=c("treatment", "evaluation_weeks", "variable")]
DT_summary <- DT_summary[,`:=`(SEMx1.96=qnorm(0.975)*norm_value_sd/sqrt(n))][, `:=`(CI.lower = norm_value-SEMx1.96, CI.upper=norm_value+SEMx1.96)]

```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#1. Evaluating assumption of normal distribution
norm_function <- function(list_object){
  if(nrow(list_object)<3){
    return(list_object[,p_value:=0])
  } else {
    return(list_object[,p_value:=shapiro.test(list_object[, norm_value])[2]])
  }
}
norm_test_p <- do.call(rbind, lapply(split(DT_melt, DT_melt[, .(treatment, evaluation_weeks, variable)]), function(subset){norm_function(subset)}))
norm_test_p <- norm_test_p[, .(p_value= mean(p_value)), by=c("treatment", "evaluation_weeks", "variable")]

#2. Evaluating assumption of homogenity of variances between treatments within time points
homo_function <- function(list_object){
  if(nrow(list_object)<3){
    return(list_object[,p_value:=0])
  } else {
    return(list_object[,p_value:=fligner.test(list_object$norm_value, list_object$treatment)[3]])
  }
}
homo_test_p <- do.call(rbind, lapply(split(DT_melt, DT_melt[, .(evaluation_weeks, variable)]), function(subset){homo_function(subset)}))
homo_test_p <- homo_test_p[, .(p_value=mean(p_value)), by=c("evaluation_weeks", "variable")]

#3. Independent two group comparison 
setkey(homo_test_p, "evaluation_weeks", "variable")
setkey(norm_test_p, "evaluation_weeks", "variable", "treatment")

two_group_test <- function(list_object){
  week = list_object[, evaluation_weeks][1]
  target = list_object[, variable][1]
  
  norm_assump_1 = FALSE
  norm_assump_2 = FALSE
  homo_assump = FALSE
  #Checking for fulfillment of assumptions
  if(homo_test_p[.(week, target), p_value]>0.05){
    homo_assump = TRUE
  }
  
  if(norm_test_p[.(week, target, "NPC"), p_value]>0.05){
    norm_assump_1 = TRUE
  }
  
  if(norm_test_p[.(week, target, "saline"), p_value]>0.05){
    norm_assump_2 = TRUE
  }
  
  if(isTRUE(norm_assump_1) & isTRUE(norm_assump_2)){
    if(isTRUE(homo_assump)){
      p_value_out <- t.test(list_object[treatment=="saline", norm_value], list_object[treatment=="NPC", norm_value], var.equal = TRUE)$p.value
      return(data.table(evaluation_week=week, variable=target, p_value=p_value_out, test="t.test_equal_var"))
    } else {
      p_value_out <- t.test(list_object[treatment=="saline", norm_value], list_object[treatment=="NPC", norm_value], var.equal = FALSE)$p.value
      return(data.table(evaluation_week=week, variable=target, p_value=p_value_out, test="t.test_unequal_var"))
    }
  } else {
    p_value_out <- wilcox.test(list_object[treatment=="saline", norm_value], list_object[treatment=="NPC", norm_value])$p.value
    return(data.table(evaluation_week=week, variable=target, p_value = p_value_out, test="wilcox_test"))
  }
}

group_comparison_p <- do.call(rbind, lapply(split(DT_melt, DT_melt[, .(variable, evaluation_weeks)]), function(subset){two_group_test(subset)}))

```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#Plotting individual targets
individual_target_plot <- function(target){
  plot_data <- DT_melt[variable==target]
  plot_data_summary <-  DT_summary[variable==target]
  p_value_data <-  group_comparison_p[variable==target]
  
  out_plot <- ggplot(plot_data_summary, aes(x=evaluation_weeks, y=norm_value, color=treatment))+
    #Main
    geom_segment(aes(x=1, xend=12, y=0, yend=0), color="black", size=1, linetype=2)+
    geom_errorbar(aes(ymin=CI.lower, ymax=CI.upper), size=3, width=1, position = position_dodge(width = 1), alpha=0.8, show.legend = FALSE)+
    geom_jitter(plot_data, mapping=aes(x=evaluation_weeks, y=norm_value, color=treatment), size=5, position = position_jitterdodge(jitter.width = 0.2, dodge.width = 1), alpha=0.7)+
    geom_point(shape = 15, size=8, position = position_dodge(width=1), show.legend = FALSE)+
    #Labels
    xlab("Weeks (Post SCI)")+
    ylab("Expression in relation to mean of healthy control")+
    scale_x_continuous(breaks=seq(0,12,1))+
    scale_y_continuous(breaks=seq(-10,20,2), limits = c(-10, 20))+
    theme(axis.title = element_text(size=16, face="bold"), legend.position = "bottom", legend.justification = "center", legend.text = element_text(size=18), legend.title = element_blank(), axis.text = element_text(size=16), axis.line.y = element_blank())+
    #Colors
    scale_color_manual(values=brewer.pal(3, "Set1"), name="Treatment:", labels=c("SCI+NPC", "SCI+Saline"))+
    #Annotations
    annotate(geom="text", label=target, x=6.5, y=7.5, size=14, fontface=2, alpha=0.8)+
    annotate(geom="text",label=paste("", toString(format(p_value_data[, p_value][1], digits=2, nsmall = 2)), sep = ""), x=2, y=18, fontface=2, size=5)+
    annotate(geom="text",label=paste("", toString(format(p_value_data[, p_value][2], digits=2, nsmall = 2)), sep = ""), x=5, y=18, fontface=2, size=5)+
    annotate(geom="text",label=paste("", toString(format(p_value_data[, p_value][3], digits=2, nsmall = 2)), sep = ""), x=12, y=18, fontface=2, size=5)+
    
    annotate(geom="text", label="UP-REGULATION", x=6.5, y=16, fontface=2, alpha=0.3, size=6)+
    annotate(geom="text", label="DOWN-REGULATION", x=6.5, y=-7, fontface=2, alpha=0.3, size=6)
  
  return(out_plot)  
}

individual_target_plot("IL-1a")

```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
################################################################### OUTPUT ##############################################################
```
## Data modifications

* *Normalisation:* Each expression value (i.e. for each day, animal and target) is divided (i.e. normalised) by the mean of the expression for the same target in healthy (none injured, none transplanted) animals.  

* *Outliers:* All observations wiht normalised expression value > 20 are removed. This cut-off is based in plotting of histogram. IL-7 is completely removed due to heavy influence of outliers. 

## Statistical analysis

###Evaluation of assumptions prior to statistical analysis
* Assumption of normality is evaluated using Shapiro Wilk's test for each treatment (group) within each time point and for each target separately. Null hypothesis is that data is normally distributed. 

* Assumption of homogenity of variances is between the two treatments (groups) within each time point and for each target. Given that both groups followed a normal distribution the homogenity of variances was evaluated using Bartlett's test. Given that at least one of the groups did not follow a normal distribution the homogenity of variances was evaluated using Fligner Killeen's test.  

###Independent intraday two group comparison
* Given that data in both treatments (groups) was normally distributed and the variances were equal the groups were compared using two-sided non-paired Student's t-test. Given that data in both treatments (groups) was normally distributed but the groups and unequal variances the variance was estimated separately for eah group and the Welch modification to the degrees of freedom was used. 

* Given that data in at least one of the groups was not normally distributed a two-sided non-paired Wilcoxon Rank Sum test with continuity correction in the normal approximation for the p-value was calculated. 


```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, fig.width=20, fig.height=10}
#INDIVIDUAL CYTOKINES
plot_grid(individual_target_plot("IL-1a"), individual_target_plot("IL-1b"), individual_target_plot("IL-2"), individual_target_plot("IL-4"))
plot_grid(individual_target_plot("IL-6"), individual_target_plot("IL-10"), individual_target_plot("IL-12(p70)"), individual_target_plot("TNF-a"))
plot_grid(individual_target_plot("IL-13"), individual_target_plot("IL-17"), individual_target_plot("IL-18"), individual_target_plot("G-CSF"))
plot_grid(individual_target_plot("GM-CSF"), individual_target_plot("GRO/KC"), individual_target_plot("IFN-g"), individual_target_plot("M-CSF"))
plot_grid(individual_target_plot("MCP-1"), individual_target_plot("MIP-1a"), individual_target_plot("MIP-3a"), individual_target_plot("RANTES"))
plot_grid(individual_target_plot("VEGF"))

```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#P-VALUES FOR INTRADAY COMPARISONS
p_table <- group_comparison_p
p_table[, "p_value"] <- p_table[, format(p_value, digits = 2)]
p_table[, "test"] = NULL
p_table <- dcast.data.table(p_table, ...~evaluation_week, value.var = "p_value")

kable(p_table, align="c", col.names = c("Target", "P-value(2w)", "P-value(5w)", "P-value(12w)"))
```

```{r echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
#SUMMARY TABLE 

```
